---
title: "FA1_EDA"
author: "MORILLO, JADE MARCO S."
date: "2025-02-02"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(qqplotr)

data <- read.csv("/Users/Dindette/Downloads/cytof_one_experiment.csv")

```

```{r }
selected_column <- "CD161"

# Density Estimate
ggplot(data, aes_string(x = selected_column)) + 
  geom_density(fill = "red", alpha = 0.5) +
  ggtitle("Density Estimate of CD161")

# ECDF Plot
ggplot(data, aes_string(x = selected_column)) + 
  stat_ecdf(geom = "step", color = "blue") + 
  ggtitle("Empirical Cumulative Distribution Function (ECDF) - CD161")
```

The density estimate plot for the CD161 column reveals that the data is right-skewed, with most values concentrated close to 0. The same can be said for the ECDF plot where it rose a lot at the beginning but the slope got less inclined after 0.

```{r }
col1 <- "KIR3DL1"
col2 <- "KIR2DS4"

# Q-Q Plot
ggplot(data, aes(sample = get(col1))) +
  stat_qq(aes(sample = get(col2)), color = "purple") +
  stat_qq_line(aes(sample = get(col2))) +
  ggtitle(paste("Q-Q Plot:", col1, "vs", col2))
```

The Q-Q plot comparing KIR3DL1 and KIR2DS4 shows that the two datasets do not follow the same distribution pattern as it deviates from the reference line.